---
title: "PML_Project"
author: "SD"
date: "Sunday, May 01, 2016"
output: html_document
---

INTRODUCTION

This goal of the project is to predict the manner in which an exercise was based on the data collected using smart devices that track the activity of 6 users.

DATA PROCESSING

Preparing the packages and libraries
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
```

Loading the data
```{r}
setwd("~/Coursera/Data_Science_Specialization/Practical_Machine_Learning/Assignment")

trainfile <- read.csv("./pml-training.csv")
dim(trainfile)
testfile  <- read.csv("./pml-testing.csv")
dim(testfile)
```

Removing columns which contain only missing values
```{r}
trainfile <- trainfile[, colSums(is.na(trainfile)) == 0] 
testfile <- testfile[, colSums(is.na(testfile)) == 0]
```

Removing unwanted columns such as the timestamps
```{r}
learn <- trainfile$classe

trainfile <- trainfile[, !grepl("^X|timestamp|window", names(trainfile))]
trainClean <- trainfile[, sapply(trainfile, is.numeric)]

trainClean$classe <- learn
dim(trainClean)

testfile <- testfile[, !grepl("^X|timestamp|window", names(testfile))]
testClean <- testfile[, sapply(testfile, is.numeric)]
dim(testClean)
```
The new dimensions of the cleaned training data are 19622 rows and 53 columns
The new dimensions of the cleaned test data are 20 rows and 53 columns

Creation of the training (70% of the file) and validation (30% of the file) data files
```{r}
set.seed(1234) 
partition <- createDataPartition(trainClean$classe, p=0.70, list=F)
DataTrain <- trainClean[partition, ]
DataValidation <- trainClean[-partition, ]
```

Plotting the correlation of the input variables in the training set
```{r}
Correlation <- cor(DataTrain[, -length(names(DataTrain))])
corrplot(Correlation, method="color")
```

DATA MODELING

Training a Random Forest model on the training data. A 5-fold cross-validation is applied.
```{r}
ConRF <- trainControl(method="cv", 5)
trainRF <- train(classe ~ ., data=DataTrain, method="rf", trControl = ConRF, ntree=250)
trainRF
```

Plotting the decision tree
```{r}
RFModel <- rpart(classe ~ ., data=DataTrain, method="class")
prp(RFModel)
```

Applying the model to the cross-validation dataset
```{r}
ValidationRF <- predict(trainRF, DataValidation)
```

Creating the confusion matrix for the cross-validation dataset
```{r}
confusionMatrix(DataValidation$classe, ValidationRF)

Accuracy <- postResample(ValidationRF, DataValidation$classe)
Accuracy

OOSError <- 1 - as.numeric(confusionMatrix(DataValidation$classe, ValidationRF)$overall[1])
OOSError
```
The Accuracy for the model on the cross-validation dataset is `r Accuracy[1]` and the Out-of-Sample Error is `r OOSError`

FINAL PREDICTION

Predicting the classe for the test data using the model used above.
```{r}
TestPrediction <- predict(trainRF, testClean[, -length(names(testClean))])
TestPrediction
```